Scaling up a GF grammar for free parsing

### 2011-12-11 doc/report/report.tex
report on gf

\section{Swedish}
\label{sec:swedish}
>> move, have GF first?\\
Swedish\cite[Inl. \textsection 3]{SAG}is a North-Germanic language,
closely related to Norwegian and Danish. The languages share most of their
grammatical structures and are mutually intelligible. Swedish is also 
an official language in Finland and altogether spoken by approximately 10 ~~
9?Check SAG million people.
Swedish syntax is often similar to English, but have some characteristics ~~:

@@

\subsection*{Verb second}
Swedish is a verb-second language\cite[p.116]{gunlog}: the
second constituent of a declarative main clause must consist of a verb.
The normal word order is subject-verb-object, but any syntactic category can be
fronted\cite[\textsection 1027]{H&H}.
Topicalisation is very common, especially for temporal and spatial adverbial phrases.
The examples \ref{ex:swedish-svo} - \ref{ex:swedish-avso} all have the same propositional
meaning, but vary in how the content is presented.
%~~ although the fronted part of sentence \ref{ex:swedish-ovs} and \ref{ex:swedish-avso} are being
%emphasized.
\enumsentence{
\shortex{4}
{Du & ser &  inte &mig.}
%empha glossning!! ha samma överallt, ' eller ''
{you & see & not & me}
{`You  don't see me'.}} \label{ex:swedish-svo}
\vspace{-3mm}
\enumsentence{
\shortexnt{4}
{Mig  & ser &  du & inte.}
{me & see & you & not}}\label{ex:swedish-ovs} 
\vspace{-3mm}
\enumsentence{
\shortexnt{4}
{Inte & ser& du &mig.}
{not&  see& you& me}} \label{ex:swedish-avso}

@@

\subsection*{Reflexive pronouns}
\label{swe:refl}
The Scandinavian language have special reflexive pronouns
and possessive pronouns for the 3rd
person\cite[\textsection 310 \& 319]{H&H}. They are distinct from the normal 3rd person forms.
\enumsentence{
\begin{tabular}{llll}
a. & Han slog \textbf{sig}.&\hspace{20mm}b. & Han såg \textbf{sitt} barn.\\
&\emph{He hit him self.} && \emph{He saw his (own) child.}
\end{tabular}}
\enumsentence{
\begin{tabular}{llll}
a. &Han slog \textbf{honom}. &b.& Han såg \textbf{hans} barn.\\
&\emph{He hit him (another person).}&&\emph{He saw his (another person's) child.}
\end{tabular}}
The 1st and 2nd persons use the normal personal pronoun in object form as reflexive
pronouns.

@@

\section{Saldo}
\label{sec:saldo}
A good parser needs a good lexicon. We have used of Saldo\cite{saldo}, a
large electronic lexicon developed and maintained at Gothenburg University. It is
built on Svenskt Associations Lexicon and contains information about more than 
120 000 modern Swedish words.
The ~~words are equipped ~~ with semantic, syntactic and a morphological
information. The user can find compounding analyses, examples of usage in corpora,
and graphs of semantically connected words.\\
\begin{figure}[h!]
\centering
\subfloat[Morphological info for ``katt"]{\label{pic:saldoTab}\includegraphics[width=40mm]{saldotab.png}}
\hspace{20mm}
\subfloat[Graph showing the hyponyms of ``katt"]{\label{pic:saldoWheel}\includegraphics[width=70mm]{saldograph.png}}
\caption{Information in Saldo}
\label{fig:saldo}
\end{figure}

@@

\section{Grammatical Framework}
\label{sec:gf}
The main component of the the project is the Grammatical Framework (GF) \cite{gfbok}, %It is
a grammar formalism based on Martin-Löf type theory\cite{martinlof}. GF is
designed for multilingual applications and the formalism is equivalent to
Parallel Multiple Context-Free Grammars (Seki \& al. 1991), it is hence
stronger than mildly contex-free grammars and can express whatever tree
adjoining grammars (Joshni,xx) and HG?? can.  --very confusing here!!

GF is a strongly typed functional programming languages, inspired by
%programming languages like 
ML \cite{ml} and Haskell \cite{haskell}. Its type
checking system is inspired by \textlambda Prolog\cite{bratko} and the
dependent types by Agda\cite{agda}. 
>> check this!!
The first version of the framework was implemented in 19xx in xx and has
since then fatt en trogen skara användare, med headquarter 
at Gothenburg University.

%> dependent and implicit types : agda. (read gf book and krasses book)
A grammar written in GF can be used for both parsing and generation.
The parsing algorithm is incremental and has polynomial time and space
complexity \cite{gfMech}. >>check!
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system.
The grammars can be compiled into portable grammar format (PGF)\ref{pgf},
which can be used with %There are libraries for using PGF files with
Java, Java script, C and Haskell. 
%>> Haskell only complete, also or C\# and C.  %may also be used in Android applications.
The interpreter, the GF shell, allows
the user to try and test her grammar by using commands for parsing, vizualization of parse trees,
random generation, word alignment, morphological quizzes etc.
%The following example shows how to parse a sentence and visualizing the parse tree.
%, by using the \textit{pipe} \verb-|-, sending the result to 
\begin{figure}[h]
\begin{verbatim} 
> parse "jag ser dig" | visualize_parse
> parse "jag ser dig" | visualize_tree
\end{verbatim}
\caption{Example of how the GF shell can be used. The first command will parse the given
sentence and visualize the resulting parse tree and the second will visualize the 
abstract syntax tree.}
\label{fig:shellvp}
\end{figure}
>> test vp -format=pdf -view=evince
>> test vt -format=pdf -view=evince -nofun

The visualization of the trees is done with graphviz\footnote{http://www.graphviz.org/}.
Figure \ref{fig:gftree1} shows the results from running the commands in figure \ref{fig:shellvp}
%a visualization of a abstract syntax tree and a parse tree.
\begin{figure}[h]
>> example, typ : i see you. 
\caption{Visualized GF trees}
\label{fig:gftree1}
\end{figure}
The reader should keep in mind that the visualized parse trees is not a
complete representation.
%>> how parse trees work, not correct complete representation.
%>> not a constituent structure. compare to lfg, fstructure, cstructure.
%>> explain how they are created, graphviz and how a user uses them.
%>> fotnoote that they are not complete

@@

>> move much of this to Swedish GF part?
Figure \ref{fig:gfSweCnc1} shows how the abstract grammar can be implemented
for Swedish. Nouns, verbs and sentences are all defined as strings, \verb|Str|,
and \verb|Pred| simply glues the two strings ``katten" and ``sover" together:\\
\verb|Pred cat sleep = "katten sover"|.\\
We get a more complicated example if we allow the nouns to be used in both
plural and singular. To do this, we add a category \verb|NP| to the
>> reformulate so that I don't sound stupid
abstract. \verb|N| now means a noun
in any of the number, and we introduce two function, \verb|NSg : N -> NP| and
\verb|NPl : N -> NP| which sets the number.
\begin{figure}[h!]
\begin{verbatim}              
 abstract TestGrammar = {          concrete TestGrammarSwe of TestGrammar = {
  cat N ; NP ; V ; S ;               lincat V, S, NP = Str ;
                                            N = {s : Num => Str} ;
  fun                                lin   
    Pred : NP -> V -> S ;              Pred n v = n ++ v ;
    NSg : N -> NP ;                    NPl n = n.s ! Pl ;
    NPl : N -> NP ;                    NSg n = n.s ! Sg ;
    cat_N : N ;                        cat_N = {s = table {Sg => "katten" ;
    sleep_V : V ;                                          Pl => "katterna"}};
 }                                     sleep_V = "sover"  ;
                                     param Num = Sg | Pl ;
                                     }
\end{verbatim}           
\caption{A modified grammar}
\label{fig:gfTest2}
\end{figure}
Figure \ref{fig:gfTest2} introduces some new concepts: records,tables and parameters.
In the concrete syntax, \verb|N| is defined to be a record consisting of the field
\verb|s|. The type of \verb|s| shows that it is a table, which given a parameter of type \verb|Num| returns
a string. \verb|Num| is defined as a parameter, which can either have value \verb|Sg|
or \verb|Pl|. In \verb|NPl| and \verb|NSg|, \verb|n.s| means that we use the
field \verb|s| of \verb|n|, and then the selection operator \verb|!| is used to
select a branch in the table.

@@

>>not very restricted, not few ambiguities.
>>the scope is controlled language, but possible ambiguities (discuss this in future
>>work, saldo, conclusion). Ambiguities: from multilinguality, or inherited:
>>syntactical (man with the telescope), 
>>lexical (youSg/Pl, växa).
A GF grammar usually covers a very restricted set of language where few ambiguities
occur and much care can be taken to preserving the semantics of the
different languages.
By using a controlled language like this,  good and reliable translation can be
ensured. Some examples of projects using GF are WebAlt\cite{webalt},
a project which aims to develop language independent material for mathematics problems,
and the formal software verification tool KeY\cite{key}, in wich GF is used for
translating specification to natural language.
%>> key, talk, hats, bom (see references in Ramonas thesis).
>> use talk here and move key to other place??
%, the Online Syllogism Solver\cite{malin!}. 
The framework is also used in the
European project MOLTO\cite{molto} for online translation between up to 15 language.\\

\subsection{The resource library}
%>>Best part? Biggest library?
%>>Most general resource, first large grammar, covers fundamental syntactics.
The GF package provides a powerful resource library\cite{gf-resource}, which covers the
fundamental morphology and syntactics for more than 20 languages. %is a very useful part of the framework. It is provided in the 
%, they give a good start for all who want to write an application grammar.
%of a language for each application, all
%this can be found as libraries. So far there are 20 languages in the library, and
The languages all share the same abstract and they constitute  
a very useful resource when implementing application grammars.
There is also a small test lexicon included, containing some hundred of common
words.

@@

A GF parse tree has two visualized representations, the \textit{parse tree} and
the \textit{abstract tree}, as shown in figure \ref{fig:gfPandAtrees}.
>> have other treees! One in english.
\begin{figure}[h]
\centering
\subfloat[GF parse tree]{\label{pic:gfPtree}\includegraphics[width=50mm]{gfPTree.pdf}}
\subfloat[GF abstract tree]{\label{pic:gfAtree}\includegraphics[width=60mm]{gfTree.pdf}}
\caption{Parse tree and abstract tree for the sentence ``Katten på bilen blir större".}
\label{fig:gfPandAtrees}
\end{figure}
The abstract trees shows all functions GF needs to use
to parse the sentence and is consequently the same for all concrete syntaxes.
The parse tree (\ref{pic:gftree}) shows the words and which types the functions assign
to them.
The abstract tree shows us that the function
\verb_DetCN_ is applied the \verb-UseN cat_N- and the information
definitive (\verb-DefArt-) and singular (\verb-NumSg-). In the Swedish parse tree
we see that \emph{``katten"}, which is in definite form singular, is noun \verb-N-
which is turned into a \verb-CN- and finally a \verb-NP-. In the English parse
tree, the form of \emph{``cat"} expresses singular while the definite article
\emph{``the"} is explicit and therefore shown in the parse tree.

\vspace{5mm}
As an open source-project, GF is constantly being developed and improved. New
languages are added, the compiler is being improved, ways of using it in more 
effecient and easier ways are added
and the possibilities to use GF in different environments
increased. There are research on how to make better use of the dependent
types, for using ontologies\cite{ramona} or generating natural
language via Montague semantics\cite{montague}.
Further, experiments on using GF for free parsing has been conducted
\cite[\textsection 4.1]{gfMech}.
This is also a future goal for our project and one of the subgoal has been to
model an important part of Swedish in GF. The starting point was the resource
grammar and the to which the result has a strong connection.

@@

\section{Talbanken}
\label{sec:talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
Being a freely available, manually annotated, large scale treebank,
Talbanken\cite{talbanken} was perfect for our purpose.
It was created in the 1970s at Lund University. In total it consists of four
parts. Two of them are transcribitions of spoken language, one a collection of
text written by high school students, and one, section P,
consists of professionally written Swedish gathered from newspapers, brochures and textbooks.\\
Talbanken was also used to train the Swedish version of the Maltparser\cite{malt}
and was thus redistributed in a more modern version\cite{talbanken05}.
%This work was done by Nivre, Hall and Nilsson in 2005.
It is released in Malt and Tiger
XML-formats\footnote{http://www.ims.uni-stuttgart.de/projekte/TIGER/TIGERSearch/doc/html/TigerXML.html},
where the trees had been made deeper and more detailed. \\
In this project, as well as for the Malt parser, the more than 6000 sentences
in section P has been used.
The treebank has served as an inspiration and an evaluation source throughout this
project. An automatic mapping between its trees and the abstract trees from GF has been
done, which will be explained in section \ref{sec:Mapping}.
%Differences in analyse, do we want a similar? 
%Uses of Talbanken in project, mapping to evaluate and test.  
%Look at output from Maltparser, could also comparing our parse trees to this.

\section{Related work}
\label{sec:related}
>> Much more here, important!! comparisons.
>> HPSG and LFG are context sensitive
Computational ling is active area for Swedish, there are many other Swedish parser,
but grammars?
Many of the parsers not ruled based, the most known one Malt\cite{malt} ... .
However has been two, about CLE\cite{cle} and what happened to it. Logic and translation
to Swedish, S-CLE (Gambäck,1997) full analyse and semantics, translation,
but coverage is too limited (nivres papper), The
spoken lang. translator\cite{cle2}. Wiren \cite{wiren}?? chartparser, but for swedish. bit more information than
just pos: cassSwe \cite{casswe}
Other Sw. parsers and grammars, evaluation of them? 
Other grammar frameworks.
SweCG (birn) annotation, parser Swedish FDG (voultanien,2001) which is not freely available.
finding subject, objects, suubject complements (SC) and
to link them to their proper regents (main verbs) : over 90\% both precision and recall.
uses a dependency grammar, but "The functional description of adverb phrases
and prepositional phrases (e.g. agent, source, goal, benefactive, time) remains
to be described in a future version." 
HPSG\cite{hpsg} and lingo matrix\cite{lingomatrix}, English, Japanese, German,
Greece, Spanish, Norweigan (NorSource, Hellan 2003),( not famous: SweCore
(Ahrenberg) based on NorSource)
also Swedish: BiTSE\cite{stymne}, Stymne 2006, Swedish English translation,
and one for the common parts of scandinavian (Haugereid etc) in Nordisk sprog.
focusing on verb phrames, built on HPSG.
S\o rgaard and Haugereid, Scandinavian Grammar Matrix\cite{scandmatrix}
Tag, Tree adjoining grammar, XTAG: English and Korean, Chinese? English: ongoing to get widecoverage grammar
under GPL. Translation to Korean.
% related and references work: CLE, Krasimirs book, CassSwe, GF book,SAG, gunlög,
% the multilingual treebank, Homse&Hinchcliff, Aarne i LILT

@@

%
%There are a number of differences between and saldo Till skillnad från GF,
%saldo does not contain forms that aren't used gf generates all forms. A form
%needed to generate the gf table may not be given in saldo such as singular
%forms for 'byxor'. Also contains idioms, 'hålla huvudet kallt' and has
%different grammatical analyse, a lot of pronouns. Example. 
One reason for why words of the categories listed above could not be imported
is that Saldo does not contain information about ~~word forms that are not used~~.
In some cases those forms are needed by the smart paradigm. 
This was a common reason why nouns could not be imported. 
Consider for example
the plural tantum %\cite[44]{SAG}
noun ``glasögon" (``glasses"). Since the smart paradigm
requires a singular form, this lemma could not be added to the lexicon. ``Glasögon" could
be added manually, by using the ~~somewhat incorrect,annan betydelse~~ singular form ``glasöga" (``glass-eye").
The same problem occurred for verbs that always are used in passive form:
>> not passive!! deponent, s-verbs!!
``andas" (``breathe"), ``frodas" (``flourish"), ``lyckas" (``succeeed"). This
was the case for close to 80\% of the failing verbs of type \verb_vb_.

@@

For verbs, but also others, there is an important distinction for valency.
Encoded in the lexicon. Categories as V,V2,VS...
Either directly to VP if V, or needs to be given arguments. Can then add adverbs.
%Maybe talk about ställa.
Particles and prepositions, can't be seen on the trees. Can now see why the parse
trees are not complete. difference shows when fronting.
tycka om /tycka om -> om vad tycker du inte. fast bra
Swedish verb may have up to five arguments according to Stymne
'jag tar med mig den till Bo' : ta\_med\_V3 = dirV3 (reflV (partV (take\_V "med")) (mkPrep "till") ;
but is really 'till' part of verb?
VP is divided corresponding to satsschema (diderichsen) (renewed by SAG).
If the object is not given immediatly, the grammar uses slash.
"Speed up", instead of movement used slashes\cite{gazdar}.
Slashes good for question (vilken katt ser du idag)
relatives (katten som du tittar på) and topicalisation
The grammar has VP,Cl, SSlash, all missing NP.

@@

\enumsentence{\shortex{3}
{Han & bara & log }
{he & only & smiles }
{`He just smiles'}}\\

@@

\subsection{Probabilities}
\label{sec:futureProbabilities}
%Getting and using probabilities, to disambiguate.
%process list from saldo.
%Use multiligual treebank?
%stop ambiguities by fixing so that Gen of a quant isn't a quant, or better, that all fields
%have a genitive field.
%More fields in the grammar? Or leave it all to the robust parser

>>>>

\section{Swedish}
\label{sec:swedish}
>> move, have GF first?\\
Swedish \cite[Inl. \textsection 3]{SAG}is a North-Germanic language,
closely related to Norwegian and Danish. The languages share most of their
grammatical structures and are mutually intelligible. Swedish is also 
an official language in Finland and altogether spoken by approximately 10 ~~
9?Check SAG million people.
Swedish syntax is often similar to English, but have some characteristics ~~:

@@

\subsection*{Verb second}
Swedish is a verb-second language \cite[p.116]{gunlog}: the
second constituent of a declarative main clause must consist of a verb.
The normal word order is subject-verb-object, but any syntactic category can be
fronted \cite[\textsection 1027]{H&H}.
Topicalisation is very common, especially for temporal and spatial adverbial phrases.
The examples \ref{ex:swedish-svo} - \ref{ex:swedish-avso} all have the same propositional
meaning, but vary in how the content is presented.
%~~ although the fronted part of sentence \ref{ex:swedish-ovs} and \ref{ex:swedish-avso} are being
%emphasized.
\enumsentence{
\shortex{4}
{Du & ser &  inte &mig.}
%empha glossning!! ha samma överallt, ' eller ''
{you & see & not & me}
{`You  don't see me'.}} \label{ex:swedish-svo}
\vspace{-3mm}
\enumsentence{
\shortexnt{4}
{Mig  & ser &  du & inte.}
{me & see & you & not}}\label{ex:swedish-ovs} 
\vspace{-3mm}
\enumsentence{
\shortexnt{4}
{Inte & ser& du &mig.}
{not&  see& you& me}} \label{ex:swedish-avso}

@@

\subsection*{Reflexive pronouns}
\label{swe:refl}
The Scandinavian language have special reflexive pronouns
and possessive pronouns for the 3rd
person \cite[\textsection 310 \& 319]{H&H}. They are distinct from the normal 3rd person forms.
\enumsentence{
\begin{tabular}{llll}
a. & Han slog \textbf{sig}.&\hspace{20mm}b. & Han såg \textbf{sitt} barn.\\
&\emph{He hit him self.} && \emph{He saw his (own) child.}
\end{tabular}}
\enumsentence{
\begin{tabular}{llll}
a. &Han slog \textbf{honom}. &b.& Han såg \textbf{hans} barn.\\
&\emph{He hit him (another person).}&&\emph{He saw his (another person's) child.}
\end{tabular}}
The 1st and 2nd persons use the normal personal pronoun in object form as reflexive
pronouns.

@@

\section{Saldo}
\label{sec:saldo}
A good parser needs a good lexicon. We have used of Saldo \cite{saldo}, a
large electronic lexicon developed and maintained at Gothenburg University. It is
built on Svenskt Associations Lexicon and contains information about more than 
120 000 modern Swedish words.
The ~~words are equipped ~~ with semantic, syntactic and a morphological
information. The user can find compounding analyses, examples of usage in corpora,
and graphs of semantically connected words.\\
\begin{figure}[h!]
\centering
\subfloat[Morphological info for ``katt"]{\label{pic:saldoTab}\includegraphics[width=40mm]{saldotab.png}}
\hspace{20mm}
\subfloat[Graph showing the hyponyms of ``katt"]{\label{pic:saldoWheel}\includegraphics[width=70mm]{saldograph.png}}
\caption{Information in Saldo}
\label{fig:saldo}
\end{figure}

@@

\section{Grammatical Framework}
\label{sec:gf}
The main component of the the project is the Grammatical Framework (GF) \cite{gfbok}, %It is
a grammar formalism based on Martin-Löf type theory \cite{martinlof}. GF is
designed for multilingual applications and the formalism is equivalent to
Parallel Multiple Context-Free Grammars (Seki \& al. 1991), it is hence
stronger than mildly contex-free grammars and can express whatever tree
adjoining grammars (Joshni,xx) and HG?? can.  --very confusing here!!

GF is a strongly typed functional programming languages, inspired by
%programming languages like 
ML \cite{ml} and Haskell \cite{haskell}. Its type
checking system is inspired by \textlambda Prolog \cite{bratko} and the
dependent types by Agda \cite{agda}. 
>> check this!!
The first version of the framework was implemented in 19xx in xx and has
since then fatt en trogen skara användare, med headquarter 
at Gothenburg University.

%> dependent and implicit types : agda. (read gf book and krasses book)
A grammar written in GF can be used for both parsing and generation.
The parsing algorithm is incremental and has polynomial time and space
complexity \cite{gfMech}. >>check!
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system.
The grammars can be compiled into portable grammar format (PGF)\ref{pgf},
which can be used with %There are libraries for using PGF files with
Java, Java script, C and Haskell. 
%>> Haskell only complete, also or C\# and C.  %may also be used in Android applications.
The interpreter, the GF shell, allows
the user to try and test her grammar by using commands for parsing, vizualization of parse trees,
random generation, word alignment, morphological quizzes etc.
%The following example shows how to parse a sentence and visualizing the parse tree.
%, by using the \textit{pipe} \verb-|-, sending the result to 
\begin{figure}[h]
\begin{verbatim} 
> parse "jag ser katten" | visualize_parse
> parse "jag ser katten" | visualize_tree
\end{verbatim}
\caption{Example of how the GF shell can be used. The first command will parse the given
sentence and visualize the resulting parse tree and the second will visualize the 
abstract syntax tree.}
\label{fig:shellvp}
\end{figure}
>> test vp -format=pdf -view=evince
>> test vt -format=pdf -view=evince -nofun

The visualization of the trees is done with graphviz\footnote{http://www.graphviz.org/}.
Figure \ref{fig:gftree1} shows the results from running the commands in figure \ref{fig:shellvp}
A GF parse tree has two visualized representations, the \textit{parse tree} and
the \textit{abstract tree}.
\begin{figure}[h]
\centering
\subfloat[GF abstract tree]{\label{pic:gfAtree}\includegraphics[width=60mm]{gfTree.pdf}}
\subfloat[GF parse tree]{\label{pic:gfStree}\includegraphics[width=50mm]{gfPTree.pdf}}
\subfloat[GF parse tree]{\label{pic:gfEtree}\includegraphics[width=50mm]{gfETree.pdf}}
\caption{Parse tree and abstract tree for the sentence ``Jag ser katten".}
\label{fig:gftree1}
\end{figure}
The abstract trees shows all functions GF needs to use
to parse the sentence.
% do not know about abstract/concrete yet: and is consequently the same for all concrete syntaxes.
The parse tree (\ref{pic:gfStree}) shows the words and which types the functions assign
to them.
The reader should keep in mind that the visualized parse trees is not a
complete representation.
If the end node of a  branch has no corresponding word, it is excluded from
the visualized parse tree. This is the case for the definite article in Swedish,
and the information \verb-DetQuant DefArt NumSg- is not shown in figure \ref{pic:gfStree}.
In the corresponding English parse, figure \ref{pic:gfEtree},
the article \emph{``the"} is explicit and therefore shown in the parse tree.

%>> how parse trees work, not correct complete representation.
%>> not a constituent structure. compare to lfg, fstructure, cstructure.

%>>the scope is controlled language, but possible ambiguities (discuss this in future
%>>work, saldo, conclusion). Ambiguities: from multilinguality, or inherited:
%>>syntactical (man with the telescope), lexical (youSg/Pl, växa).
%>> have ambiguity section. 
The scope of GF grammars is usually controlled language. By
restricting the coverage, the number of ambiguities
can be limited and we can ensure that the semantics is preserved during
translation. The ambiguities arising from multilinguality and the inherited will of
course remain, but limited. %Example of domainspecific word?
The use of controlled language gives a possibility of good and reliable translation.
Some examples of projects using GF are WebAlt \cite{webalt},
a project which aims to develop language independent material for mathematics problems,
and the formal software verification tool KeY \cite{key}, in wich GF is used for
translating specification to natural language.
%>> key, talk, hats, bom (see references in Ramonas thesis).
%>> use talk here and move key to other place??
The framework is also used in the
European project MOLTO \cite{molto} for online translation between up to 15 language.\\

@@

%>> move much of this to Swedish GF part?
Figure \ref{fig:gfSweCnc1} shows how the abstract grammar can be implemented
for Swedish. Nouns, verbs and sentences are all defined as strings, \verb|Str|,
and \verb|Pred| simply glues the two strings ``katten" and ``sover" together:\\
\verb|Pred cat sleep = "katten sover"|.\\
We get a more complicated example if we allow the nouns to be used in both
plural and singular. We add a category \verb|NP| to the
abstract. Note that \verb-NP- is normally used for nouns with a fixed definitness, but for 
now we concentrate on the number only. Hence \verb|N| now means a noun
in any of the number, and we introduce two function, \verb|NSg : N -> NP| and
\verb|NPl : N -> NP| which sets the number.
\begin{figure}[h!]
\begin{verbatim}              
 abstract TestGrammar = {          concrete TestGrammarSwe of TestGrammar = {
  cat N ; NP ; V ; S ;               lincat V, S, NP = Str ;
                                            N = {s : Num => Str} ;
  fun                                lin   
    Pred : NP -> V -> S ;              Pred n v = n ++ v ;
    NSg : N -> NP ;                    NPl n = n.s ! Pl ;
    NPl : N -> NP ;                    NSg n = n.s ! Sg ;
    cat_N : N ;                        cat_N = {s = table {Sg => "katten" ;
    sleep_V : V ;                                          Pl => "katterna"}};
 }                                     sleep_V = "sover"  ;
                                     param Num = Sg | Pl ;
                                     }
\end{verbatim}           
\caption{A modified grammar}
\label{fig:gfTest2}
\end{figure}
Figure \ref{fig:gfTest2} introduces some new concepts: records,tables and parameters.
In the concrete syntax, \verb|N| is defined to be a record consisting of the field
\verb|s|. The type of \verb|s| shows that it is a table, which given a
parameter of type \verb|Num| returns a string. \verb|Num| is defined as a
parameter, which can either have value \verb|Sg|
or \verb|Pl|. In \verb|NPl| and \verb|NSg|, \verb|n.s| means that we use the
field \verb|s| of \verb|n|, and then the selection operator \verb|!| is used to
select a branch in the table.

@@

\subsection{The resource library}
%>>Best part? Biggest library?
%>>Most general resource, first large grammar, covers fundamental syntactics.
The GF package provides a powerful resource library \cite{gf-resource}, which covers the
fundamental morphology and syntactics for more than 20 languages. %is a very useful part of the framework. It is provided in the 
%, they give a good start for all who want to write an application grammar.
%of a language for each application, all
%this can be found as libraries. So far there are 20 languages in the library, and
The languages all share the same abstract and they constitute  
a very useful resource when implementing application grammars.
There is also a small test lexicon included, containing some hundred of common
words.

@@

Section \ref{sec:swegf} will give a more thorough description of the Swedish resource grammar.

\subsection{Frontiers of Grammatical Framework}
As an open source-project, GF is constantly being developed and improved. New
languages are added, the compiler is being improved, ways of using it in more 
effecient and easier ways are added
and the possibilities to use GF in different environments
increased. There are research on how to make better use of the dependent
types, for using ontologies for reasoning \cite{ramona} or generating natural
language via Montague semantics \cite{montague}.
Further, experiments on using GF for free parsing has been conducted
\cite[\textsection 4.1]{gfMech}.
This is also a future goal for our project and one of the subgoal has been to
model an important part of Swedish in GF. The starting point was the resource
grammar to which the result has a strong connection.

@@

\section{Talbanken}
\label{sec:talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
Being a freely available, manually annotated, large scale treebank,
Talbanken \cite{talbanken} was perfect for our purpose.
It was created in the 1970s at Lund University. In total it consists of four
parts. Two of them are transcribitions of spoken language, one a collection of
text written by high school students, and one, section P,
consists of professionally written Swedish gathered from newspapers, brochures and textbooks.\\
Talbanken was also used to train the Swedish version of the Maltparser \cite{malt}
and was thus redistributed in a more modern version \cite{talbanken05}.
%This work was done by Nivre, Hall and Nilsson in 2005.
It is released in Malt and Tiger
XML-formats\footnote{http://www.ims.uni-stuttgart.de/projekte/TIGER/TIGERSearch/doc/html/TigerXML.html},
where the trees had been made deeper and more detailed. \\
In this project, as well as for the Malt parser, the more than 6000 sentences
in section P has been used.
The treebank has served as an inspiration and an evaluation source throughout this
project. An automatic mapping between its trees and the abstract trees from GF has been
done, which will be explained in section \ref{sec:Mapping}.
%Differences in analyse, do we want a similar? 
%Uses of Talbanken in project, mapping to evaluate and test.  
%Look at output from Maltparser, could also comparing our parse trees to this.

\section{Related work}
\label{sec:related}
>> Much more here, important!! comparisons.
>> HPSG and LFG are context sensitive
Computational ling is active area for Swedish, there are many other Swedish parser,
but grammars?
Many of the parsers not ruled based, the most known one Malt \cite{malt} ... .
However has been two, about CLE \cite{cle} and what happened to it. Logic and translation
to Swedish, S-CLE (Gambäck,1997) full analyse and semantics, translation,
but coverage is too limited (nivres papper), The
spoken lang. translator \cite{cle2}. Wiren \cite{wiren}?? chartparser, but for swedish. bit more information than
just pos: cassSwe \cite{casswe}
Other Sw. parsers and grammars, evaluation of them? 
Other grammar frameworks.
SweCG (birn) annotation, parser Swedish FDG (voultanien,2001) which is not freely available.
finding subject, objects, suubject complements (SC) and
to link them to their proper regents (main verbs) : over 90\% both precision and recall.
uses a dependency grammar, but "The functional description of adverb phrases
and prepositional phrases (e.g. agent, source, goal, benefactive, time) remains
to be described in a future version." 
HPSG \cite{hpsg} and lingo matrix \cite{lingomatrix}, English, Japanese, German,
Greece, Spanish, Norweigan (NorSource, Hellan 2003),( not famous: SweCore
(Ahrenberg) based on NorSource)
also Swedish: BiTSE \cite{stymne}, Stymne 2006, Swedish English translation,
and one for the common parts of scandinavian (Haugereid etc) in Nordisk sprog.
focusing on verb phrames, built on HPSG.
S\o rgaard and Haugereid, Scandinavian Grammar Matrix \cite{scandmatrix}
Tag, Tree adjoining grammar, XTAG: English and Korean, Chinese? English: ongoing to get widecoverage grammar
under GPL. Translation to Korean.
% related and references work: CLE, Krasimirs book, CassSwe, GF book,SAG, gunlög,
% the multilingual treebank, Homse&Hinchcliff, Aarne i LILT

@@

%
%There are a number of differences between and saldo Till skillnad från GF,
%saldo does not contain forms that aren't used gf generates all forms. A form
%needed to generate the gf table may not be given in saldo such as singular
%forms for 'byxor'. Also contains idioms, 'hålla huvudet kallt' and has
%different grammatical analyse, a lot of pronouns. Example. 
One reason for why words of the categories listed above could not be imported
is that Saldo does not contain information about ~~word forms that are not used~~.
In some cases those forms are needed by the smart paradigm. 
This was a common reason why nouns could not be imported. 
Consider for example
the plural tantum % \cite[44]{SAG}
noun ``glasögon" (``glasses"). Since the smart paradigm
requires a singular form, this lemma could not be added to the lexicon. ``Glasögon" could
be added manually, by using the ~~somewhat incorrect,annan betydelse~~ singular form ``glasöga" (``glass-eye").
The same problem occurred for verbs that always are used in passive form:
>> not passive!! deponent, s-verbs!!
``andas" (``breathe"), ``frodas" (``flourish"), ``lyckas" (``succeeed"). This
was the case for close to 80\% of the failing verbs of type \verb_vb_.

@@

For verbs, but also others, there is an important distinction for valency.
Encoded in the lexicon. Categories as V,V2,VS...
Either directly to VP if V, or needs to be given arguments. Can then add adverbs.
%Maybe talk about ställa.
Particles and prepositions, can't be seen on the trees. Can now see why the parse
trees are not complete. difference shows when fronting.
tycka om /tycka om -> om vad tycker du inte. fast bra
Swedish verb may have up to five arguments according to Stymne
'jag tar med mig den till Bo' : ta\_med\_V3 = dirV3 (reflV (partV (take\_V "med")) (mkPrep "till") ;
but is really 'till' part of verb?
VP is divided corresponding to satsschema (diderichsen) (renewed by SAG).
If the object is not given immediatly, the grammar uses slash.
"Speed up", instead of movement used slashes \cite{gazdar}.
Slashes good for question (vilken katt ser du idag)
relatives (katten som du tittar på) and topicalisation
The grammar has VP,Cl, SSlash, all missing NP.

@@

\enumsentence{\shortex{3}
{Han & bara & log }
{he & only & smiles }
{`He just smiles'}}

@@

\subsection{Probabilities}
\label{sec:futureProbabilities}
% discuss ambiguites, will always have some and the big lexicon will increase them.
% exapmle 'vad händer under äktenskapet?'
%Getting and using probabilities, to disambiguate.
%process list from saldo.
%Use multiligual treebank?
%stop ambiguities by fixing so that Gen of a quant isn't a quant, or better, that all fields
%have a genitive field.
%More fields in the grammar? Or leave it all to the robust parser
